---
title: "Log file of analysis breathing peak flow"
author: "Daniël Pastoor"
output:
  html_document:
    toc: true
    theme: united
---


# Preliminaries

This research was conducted during the period from the end of April to the end of June 2023. This was done mainly on the Zernike campus in Groningen. With an exception of the korfball club in Zwolle.


# Introduction

## Goal
> The goal is to find a relation between sporting and non sporting poeples in connection with there breathing peak flow before and after an excersise. 

## Data collection
People data were collected mostly at Hanze University on the Zernike Campas. But there was also data collected from a korfball club in utrecht. The data was collected on the prefered location of the subject. Then we began with measuring his breathing peak flow. Following with 10 jumping jacks and measured it again.

## Theory
Background information from literature!!

- What is agentX
- What is adrenaline & how is it determined?
- etc.

# Experimental work

## April 17, 2023

## May 3, 2023  

## May 10, 2023

## May 11, 2023

1. Been busy today testing Peak flow meter. To see if it still needed to be calibrated.
2. after testing we were able to get some data from our class 
3. Furthermore I went with my group towards willemdewijk school to find test persons.
4. Also after that we went to ACLO to research more people there"

## May 12, 2023

1. Been busy today adjusting long capasity to flow peak.
2. Further, today I created and set up the Git Repository with the proper directory structure
3. In the process also created and completed the readme

## May 15, 2023

### Data reading

Data were read from file:

```{r}
data <- read.table(file = "../raw_data/Data_breath.csv", header = T, sep = ",")
dim(data)
head(data)
```
This seems OK. The data dimensions and data types are according to what was expected.

### Test of normality


Today I am going to determine whether there is a significant difference between the before and after measurements.

I want to carry out a t-test.

To do that, I need to assess if the samples are normally distributed.

This is the first analysis.

```{r test-of-normality}
shapiro.test(data$Before)
shapiro.test(data$After)
```
There is no reason to assume non-normality here for the two data columns.

## T-test

```{r}
with(data, 
    t.test(x = Before, 
           y = After,
           paired = TRUE)
)
```
The test indicates there is a highly significant difference in Before and After


## Playground

