---
title: "Log file of analysis breathing peak flow"
author: "Daniël Pastoor"
output:
  html_document:
    toc: true
    theme: united
---


# Preliminaries

This research was conducted during the period from the end of April to the end of June 2023. This was done mainly on the Zernike campus in Groningen. With an exception of the korfball club in Zwolle.


# Introduction

## Goal
> The goal is to find a relation between sporting and non sporting poeples in connection with there breathing peak flow before and after an excersise. 

## Data collection
People data were collected mostly at Hanze University on the Zernike Campas. But there was also data collected from a korfball club in utrecht. The data was collected on the prefered location of the subject. Then we began with measuring his breathing peak flow. Following with 10 jumping jacks and measured it again.

## Theory
The purpose of this experiment was to determine whether or not sports/working out influences the peak flow rate of students. The peak flow rate is the maximum rate of airflow that a person can exhale and is measured in liters per minute.  A healthy peak flow rate lies between 400 and 600 liters per minute. A peak flow rate between 200 and 400 liters per minute can indicate asthma. The peak flow rate is measured using a spirometer. 

Previous research has proven that exercising can increase one's peak flow rate. In this experiment an attempt to prove this will be done by measuring the peak flow rates of subjects before and after performing 10 jumping jacks. The subjects will be categorized in subjects that participate in sports/work out and subjects that don’t. 

# Experimental work

## April 17, 2023

Today we did a before invesigation about what kind for subject we want to analyse.

## May 10, 2023

1. Today we started with our first testing on the zernike cammpus

## May 11, 2023

1. Been busy today testing Peak flow meter. To see if it still needed to be calibrated.
2. after testing we were able to get some data from our class 
3. Furthermore I went with my group towards willemdewijk school to find test persons.
4. Also after that we went to ACLO to research more people there"

## May 12, 2023

1. Been busy today adjusting long capasity to flow peak.
2. Further, today I created and set up the Git Repository with the proper directory structure
3. In the process also created and completed the readme

## May 15, 2023

1. Today we started testing on te zernike campus in the Van Doornveste building. With the target to find no sporting subjects

## May 17, 2023

1. Today I started with creating the R markdown file and started with filling in some data
2. Also I began with reading the data

### Data reading

Data were read from file:

```{r}
data <- read.table(file = "../raw_data/Data_breath.csv", header = T, sep = ",")
dim(data)
head(data)
```
This seems OK. The data dimensions and data types are according to what was expected.

## May 22, 2023

1. Today I added an ignore file so the data is not stored in the git repository.
2. Also I added the hypothesis with the correct introduction.
3. I added the protocol
4. Changed some translation from dutch to english
5. We sorted the data and changed the ; to ,

## May 23, 2023

1. Today we created a planner for who will do what.
2. I added my 4 histograms about the diffrent groups.

```{r histograms}
# histogram of man that sport
ggplot(data = data_breath %>% filter(Sex == "Man")) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Sporting), bins = 30, position="dodge", right=TRUE) +
       labs(fill='Jumping Jacks',
            x = "Breahing peak flow rate in L/m",
       alt = "A histogram plotted from a data collection with the before and after breathing peak flow of man")

# histogram of womans that do not sport
ggplot(data = data_breath %>% filter(Sex == "Woman")) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Sporting), bins = 30, position="dodge", right=TRUE) +
       labs(fill='Jumping Jacks',
            x = "Breahing peak flow rate in L/m",
       alt = "A histogram plotted from a data collection with the before and after breathing peak flow of woman")

# histogram of man that sport before and after 10 jumping jacks
ggplot(data = data_breath %>% filter(Sex == "Man") %>% filter(Sporting == T)) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Has_activity), bins = 30, position="dodge", right=TRUE) +
       labs(fill='Jumping Jacks',
            x = "Breahing peak flow rate in L/m",
       alt = "A histogram plotted from a data collection with the before and after breathing peak flow of man")

# histogram of womans that do not sport
ggplot(data = data_breath %>% filter(Sex == "Woman") %>% filter(Sporting == T)) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Has_activity), bins = 30, position="dodge", right=TRUE) +
       labs(fill='Jumping Jacks',
            x = "Breahing peak flow rate in L/m",
       alt = "A histogram plotted from a data collection with the before and after breathing peak flow of man")


# histogram of man that sport before and after 10 jumping jacks
ggplot(data = data_breath %>% filter(Sex == "Man") %>% filter(Sporting == F)) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Has_activity), bins = 30, position="dodge", right=TRUE) +
       labs(fill='Jumping Jacks',
            x = "Breahing peak flow rate in L/m",
       alt = "A histogram plotted from a data collection with the before and after breathing peak flow of man")

# histogram of womans that do not sport
ggplot(data = data_breath %>% filter(Sex == "Woman") %>% filter(Sporting == F)) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Has_activity), bins = 30, position="dodge", right=TRUE) +
       labs(fill='Jumping Jacks',
            x = "Breahing peak flow rate in L/m",
       alt = "A histogram plotted from a data collection with the before and after breathing peak flow of man")
```

## 31 May, 2023

1. Todat Marian and I have created the title for our final paper.

## 1 June, 2023

1. Today I checked my Histogram and had a meeting with my group!

## 2 June, 2023

2. Today I helped larissa with reacting the nessasery Anova test for our data.

## 3 June, 2023

Today I optimized the histogram to a final histogram. I did this to better visualize the data
```{r}
# To add labels for the face_wrap
data_breath <-  data_breath %>%
  mutate( Has_activity_label = case_when(Has_activity == "True" ~ "After jumping jacks" , .default = "Before jumping jacks"))

# histogram of people that sport
ggplot(data = data_breath %>% filter(Sporting == T)) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Sex), bins = 30, position="dodge", right=TRUE) +
      facet_wrap(~ Has_activity_label) +
       labs(fill='Sex',
            x = "Breahing peak flow rate in L/m",
       alt = "A histogram plotted from a data collection with the filter if the person sport grouped by sex")

# histogram of people that do not sport
ggplot(data = data_breath %>% filter(Sporting == F)) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Sex), bins = 30, position="dodge", right=TRUE) +
  facet_wrap(~ Has_activity_label) +
       labs(fill='Sex',
            x = "Breahing peak flow rate in L/m",
       alt = "A histogram plotted from a data collection with the filter if the person does not sport grouped by sex")
```

## June 9-6

### optimized box plot:
# Marian
```{r }
females_n <- data_breath %>%
  filter(Sex == "Woman", Sport == "None")

before_jj_nf <- females_n[c(rep(TRUE,1),FALSE), ]


after_jj_nf <- females_n[c(rep(FALSE,1),TRUE), ]



ggplot(data = before_jj_nf, 
       mapping = aes(x = Sex, y = Peak_flow, color = Sex)) +
  geom_boxplot() + 
  stat_summary(fun = "mean", geom = "point",
               color = "gold") + xlab("Peak flow rate in females that don't exercise before performing 10 jumping jacks")


ggplot(data = after_jj_nf,
       mapping = aes(x = Sex, y = Peak_flow, color = Sex)) +
      geom_boxplot() +
      stat_summary(fun = "mean", geom = "point",
               color = "gold") +  xlab("Peak flow rate in females that don't exercise after performing 10 jumping jacks")


#Boxplots of peak flow rate in females that do exercise before and after jumping jacks

females_y <- data_breath %>%
  filter(Sex == "Woman", Sport != "None")

print(females_y)

before_jj_yf <- females_y[c(rep(TRUE,1),FALSE), ]


after_jj_yf <- females_y[c(rep(FALSE,1),TRUE), ]



ggplot(data = before_jj_yf,
       mapping = aes(x = Sex, y = Peak_flow, color = Sex)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point",
               color = "gold") + xlab("Peak flow rate in females that do exercise before performing 10 jumping jacks")


ggplot(data = after_jj_yf,
       mapping = aes(x = Sex, y = Peak_flow, color = Sex)) +
      geom_boxplot() +
      stat_summary(fun = "mean", geom = "point",
               color = "gold") +  xlab("Peak flow rate in females that do exercise after performing 10 jumping jacks")




#Boxplots of peak flow rate in males that don't exercise before and after jumping jacks
males_n <- data_breath %>%
  filter(Sex == "Man", Sport == "None")

before_jj_nm <- males_n[c(rep(TRUE,1),FALSE), ]


after_jj_nm <- males_n[c(rep(FALSE,1),TRUE), ]



ggplot(data = before_jj_nm,
       mapping = aes(x = Sex, y = Peak_flow, color = Sex)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point",
               color = "gold") + xlab("Peak flow rate in females that don't exercise before performing 10 jumping jacks")


ggplot(data = after_jj_nm,
       mapping = aes(x = Sex, y = Peak_flow, color = Sex)) +
      geom_boxplot() +
      stat_summary(fun = "mean", geom = "point",
               color = "gold") +  xlab("Peak flow rate in males that don't exercise after performing 10 jumping jacks")


#Boxplots of peak flow rate in males that do exercise before and after jumping jacks
males_n <- data_breath %>%
  filter(Sex == "Man", Sport != "None")

before_jj_ym <- males_n[c(rep(TRUE,1),FALSE), ]


after_jj_ym <- males_n[c(rep(FALSE,1),TRUE), ]



ggplot(data = before_jj_ym,
       mapping = aes(x = Sex, y = Peak_flow, color = Sex)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point",
               color = "gold") + xlab("Peak flow rate in females that do exercise before performing 10 jumping jacks")


ggplot(data = after_jj_ym,
       mapping = aes(x = Sex, y = Peak_flow, color = Sex)) +
      geom_boxplot() +
      stat_summary(fun = "mean", geom = "point",
               color = "gold") +  xlab("Peak flow rate in males that do exercise after performing 10 jumping jacks")
```
To


### paper
// histogram sportend en niet sported peak flow vergelijken

// maximum en minum average
// conclusie weg en signifififificsnt weg
Vanuit de verzamelde gegevens kunnen we de volgende deelvragen beantwoorden:

// er is wel een verschil 

Bestaat er een verband tussen de frequentie en het type lichaamsbeweging en de piekstroom vóór en na fysieke activiteit bij mannelijke en vrouwelijke studenten tussen de 18 en 25 jaar?
Is er een hogere piekstroomsnelheid bij mannelijke en vrouwelijke proefpersonen tussen de 18 en 25 jaar die aan lichaamsbeweging doen, in vergelijking met proefpersonen die dat niet doen?
Als we kijken naar de piekstroom vóór en na fysieke activiteit, zien we zeker een significant verschil tussen mannen en vrouwen die wel of niet sporten. Dit blijkt uit de onderstaande grafiek, waarin te zien is dat de piekstroom na 10 jumping jacks hoger is bij mensen die sporten dan bij mensen die dat niet doen. Bovendien is de ademhalingspiekstroom na 10 jumping jacks lager bij mensen die niet sporten, in vergelijking met mensen die wel sporten. Op basis hiervan kunnen we concluderen dat er zeker een verschil is in de piekstroom bij mannen en vrouwen die wel of niet sporten, met betrekking tot hun ademhalingspiekstroom.

Als we verder kijken naar de piekstroom vóór en na fysieke activiteit en dit vergelijken tussen mannen en vrouwen, kunnen we concluderen dat er geen significant verschil is. Wel is te zien dat de piekstroomsnelheid bij mannen hoger is, maar dit kan ook te maken hebben met de lichaamsbiologie van mannen en vrouwen. Als we echter kijken naar de verschillen tussen de voor- en nawaarden, zien we dat deze ongeveer gelijk zijn. Hieruit kunnen we concluderen dat het geslacht geen invloed heeft op de ademhalingspiekstroom wanneer we naar het verschil kijken. De hoogte van de waarde is echter mogelijk gerelateerd aan de lichaamsbiologie.

__

Presenteer je resultaten in een logische volgorde. Beschrijf wat er te zien is en werk toe naar het beantwoorden van je doelstelling. Beschrijf indien mogelijke de logica van de keuze voor opeenvolgende experimenten. In elke paragraaf worden de feitelijke conclusies gegeven (bv ‘de vergelijking laat zien dat het gemiddelde van groep A significant afwijkt van het gemiddelde van groep B, met een p-waarde van … Dit komt niet overeen met de in de literatuur (REF) beschreven waardes.’)
Daniel Plots en uitleg
Larissa teste
]

```{r, warning=FALSE, fig.cap='A histogram plotted from a data collection with the filter if the person sport grouped by sex'}
# To add labels for the face_wrap
data_breath <-  data_breath %>%
  mutate( Has_activity_label = case_when(Has_activity == "True" ~ "After jumping jacks" , .default = "Before jumping jacks"))

# histogram of people that sport
ggplot(data = data_breath %>% filter(Sporting == T)) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Sex), bins = 30, position="dodge", right=TRUE) +
      facet_wrap(~ Has_activity_label) +
       labs(fill='Sex',
            x = "Breahing peak flow rate in L/m")

```

```{r, warning=FALSE, fig.cap='A histogram plotted from a data collection with the filter if the person does not sport grouped by sex'}
# To add labels for the face_wrap
data_breath <-  data_breath %>%
  mutate( Has_activity_label = case_when(Has_activity == "True" ~ "After jumping jacks" , .default = "Before jumping jacks"))

# histogram of people that do not sport
ggplot(data = data_breath %>% filter(Sporting == F)) +
       geom_histogram(mapping = aes(x = Peak_flow,  fill=Sex), bins = 30, position="dodge", right=TRUE) +
  facet_wrap(~ Has_activity_label) +
       labs(fill='Sex',
            x = "Breahing peak flow rate in L/m",
       alt = "A histogram plotted from a data collection with the filter if the person does not sport grouped by sex")
```